A Graph Theory Based Approach to MDAO Problem Formulation

### 2013-04-18 graph_syntax.tex
split graph_syntax into two files

These operations, though very simple, are effectively the same kind of task 
performed by an analysis block. A constraint or objective function can therfore 
be represented as another analysis block on the graph, with it's own input and 
output variable nodes. Although fundamentally no different than an analysis block, 
for clarity and convience it is useful to distinguish between regular analysis 
blocks and those that arise from the addition of objectives or constraints to 
the graph. So we define an expression block, as the collection of variable and model 
nodes related to a given objective or constraint function. Fig. \ref{f:obj-cons}
highlights the objective and constraint blocks from the Sellar problem.

@@

\section{Building MCG and FPG}
	\label{s:building graphs}
\subsection{Maximal Connectivity Graph}
	\label{ss:MCG}
	To construct the maximal connectivity graph, we assume that a set of codes, global inputs, and objectives and constraints (collectively called global outputs) are provided. The codes are represented by analysis block graphs ((ABGs?)) $A_i=(V_{A_i},E_{A_i}), \ i\in I$, where $m$ is the number of codes and $I=\{1,2,\ldots,m\}$. The global inputs are represented as a set of variable nodes $V_\txt{in}$, and the global outputs are represented by a set of variable nodes $V_\txt{out}$. 
	We assume that $V_\txt{in}$, $V_\txt{out}$, and each $A_i$ are given, and that any potential connection between variables is given in the form of connection edges in the set $C_M$. 
	Then we may construct the maximal connectivity graph $M=(V_M,E_M)$ as
	\begin{IEEEeqnarray*}{rCl}
	V_M & = & V_\txt{in} \cup V_\txt{out} \cup \left( \bigcup_{i \in I} V_{A_i} \right), \\
	E_M & = & C_M \cup \left( \bigcup_{i \in I} E_{A_i} \right),
	\end{IEEEeqnarray*}
	The MCG $M$ is uniquely determined by the given set of analysis blocks, the required outputs, and the given global inputs. In the cases where the set of global inputs is not known a priori, the process of obtaining the FPG will reveal the required inputs, as discussed subsequently.

\subsection{Fundamental Problem Graph}
	\label{ss:FPG}
	We now define the fundamental problem graph $F=(V_F,E_F)$ as a directed graph meeting the following conditions, which are explained subsequently,
	\begin{enumerate}
	\item[(1)] $\displaystyle{I_F \subset I}$ such that the remaining conditions hold
	\item[(2)] $\displaystyle{ \forall i \in I_F, \ \exists v \in V_{A_i} \txt{ with } t_\txt{node}(E^{-1}(v))=\txt{`model'} \txt{ and } \txt{deg}^+(v)>0 }$
	%\item[(3)] $\displaystyle{ \forall i \in I, \forall v \in V_{A_i} \txt{ with } t_\txt{node}(v) = \txt{`variable,'} \  \txt{deg}_l^-(v) < \txt{deg}^-(v) \leq \txt{deg}_u^-(v)}$
	%\item[(3)] $\displaystyle{ \forall i \in I, \forall v \in V_\txt{out} \txt{ with } t_\txt{node}(v) = \txt{`variable,'} \  \txt{deg}_l^-(v) < \txt{deg}^-(v) \leq \txt{deg}_u^-(v)}$
	\item[(3)] $\displaystyle{V_{\txt{in},F} = \{v \in V_\txt{in} \st \txt{deg}^+(v) > 0\} }$ ((maybe this should be 1 instead of 0 to match the previous definition of global input))
	\item[(4)] $\displaystyle{V_F = V_{\txt{in},F} \cup V_\txt{out} \cup \left( \bigcup_{i \in I_F} V_{A_i} \right)}$
	\item[(5)] $\displaystyle{E_F = C_F \cup \left( \bigcup_{i \in I_F} E_{A_i} \right), \ C_F \subset C_M}$
	\item[(6)] $\displaystyle{\forall v \in V_F \txt{ with } t_\txt{node}(v) = \txt{`variable,'} \  \txt{deg}_l^-(v) \leq \txt{deg}^-(v) \leq \txt{deg}_u^-(v)}$
	\end{enumerate}
	The set $I_F$ is an index set containing the indices of the analysis blocks in $F$ and it may only include the analysis blocks $A_i$ that meet requirements (2) and (6). 
	Requirement (2) stipulates that each analysis blocks in $F$ must have at least one local output that is being used. 
	Requirement (3) stipulates that only the global inputs that are being used should be included. 
	Requirements (4) and (5) provide the construction of the sets of nodes and edges, respectively. 
	Lastly, requirement (6) stipulates that the number of edges directed into the variable nodes must be within the lower and upper indegree limits; if $\txt{deg}^-(v) < \txt{deg}_l^-(v)$ the node is called a \emph{hole}, and if $\txt{deg}^-(v) > \txt{deg}_u^-(v)$ the node is called a \emph{collision}.

	The set $C_F$ is a set containing connection edges representing connections. Since $C_M$ is the set of all potential connections, we must have $C_F \subset C_M$. While no other conditions are explicitly stated for $C_F$, requirement (6) is actually a requirement on both $I_F$ and $C_F$.

\subsection{Obtaining the Fundamental Problem Graph}
	\label{ss:obtaining FPG}
	In general, there may be multiple different graphs that satisfy the FPG conditions, though there may be none at all. Here, we describe a process for obtaining an FPG by starting with the MCG and disconnecting connection edges until the FPG conditions are met. Then the problem is reduced to deciding which connection edges to remove.

	%Then, mathematically, the FPG starts with $\mathcal A_1 = \{1,\ldots,m\}$, $I_{F,1} = I$, and $C_{F,1} = C_M$.
	Then, mathematically, the FPG starts with $C_{F,0} = C_M$, $E_{F,0} = E_M$, $V_{F,0} = V_M$.
	The nodes where $\txt{deg}^-(v) < \txt{deg}_l^-(v)$ are considered first.
	\begin{enumerate}
	\item The first step is to detect holes and disconnect the connection edges following them. These connection edges are removed because they represent variables which cannot be determined because the analysis function does not have adequate inputs. The set of variable nodes which are holes is created as
	\begin{equation}
	H = \{v \in V_M \st t_\txt{node}(v) = \txt{`variable'} \txt{ and } \txt{deg}^-(v) < \txt{deg}_l^-(v) \},
	\end{equation}
	which is the set of variable nodes with fewer incoming edges than are allowed by the lower indegree limit.
	Then the updated set of edges is created by removing the first set of connection edges following the hole:
	\begin{equation}
	C_{F,1} = C_{F,0} \setminus \{(x,y) \in E_M \st x=E_M^3(v) \txt{ for } v \in H\},
	\end{equation}
	%deg-v is now recaculated for the new C. 
	and this step is demonstrated by Fig.~\ref{f:hole}. Because removing these edges can create new holes, this step must be repeated until no more holes are found.
	\begin{figure}[htb!]
		\begin{center}
		\includegraphics[width=3in]{images/analysis_block_hole}
		\end{center}
		\vspace{-20pt}
	\caption{Example variable node indicating a hole.}
	\label{f:hole}
	\end{figure}

	\item The second step is to detect collisions and to then disconnect enough connection edges so that the collisions are resolved. The set of variable nodes which contain collisions is created as
	\begin{equation}
	S_\txt{nodes} = \{v \in V_M \st t_\txt{node}(v) = \txt{`variable'} \txt{ and } \txt{deg}^-(v) > \txt{deg}_u^-(v) \}.
	\end{equation}
	For each collision node we can construct a set contained the edges directed in. The set containing all of these sets is constructed as
	\begin{equation}
	S_\txt{edges} = \big \{ \{(x,y) \in E_M\} \st y \in S_\txt{nodes} \big \}
	\end{equation}
	Let $J=\{1,2,\ldots,|S_\txt{edges}|\}$ be an indexing set for $S_\txt{edges}$ such that each $S_{\txt{edges},j}$ corresponds to a set in $S_\txt{edges}$ for $j \in J$. 
	An example collision is shown in Fig.~\ref{f:collision} to indicate the definition of $S_{\txt{edges},j}$. 
	We can assume that $J$ also indexes $S_\txt{nodes}$ because there is a one--to--one correspondence between the elements in $S_\txt{nodes}$ and the elements in $S_\txt{edges}$ (which are sets). 
	Then we may construct sets of edges as
	\begin{equation}
	B_j = \big \{e_k \in S_{\txt{edges},j} \st k \in \{1,2,\ldots,K\} \txt{ with } K \leq \txt{deg}_u^-(v_j) \big \}, \ j \in J,
	\end{equation}
	which means that each set $B_j$ is constructed from the set $S_{\txt{edges},j}$ by taking only as many edges as are allowed by the upper indegree limit of $v_j$. The construction of each $B_j$ corresponds to making a decision on which edges to include and which edges not to include. Then let
	\begin{equation}
	C_{F,2} = \{ e \in C_{F,1} \st e \in B_j \txt{ for some } j \in J\},
	\end{equation}
	\begin{equation}
	E_{F,2} = E_M \setminus (C_M \setminus C_{F,2})
	\end{equation}
	and
	\begin{equation}
	F_2 = (V_M,E_{F,2})
	\end{equation}
	\begin{figure}[htb!]
		\begin{center}
		\includegraphics[width=3in]{images/analysis_block_collision}
		\end{center}
		\vspace{-20pt}
	\caption{Example variable node indicating a collision.}
	\label{f:collision}
	\end{figure}

	\item The third and final step is to remove any unused nodes and edges. The indexing set for only the analysis blocks with at least one used local output may be constructed as ((pruning))
	\begin{equation}
	I_F =  \{i \in I \ | \ \exists v \in V_{A_i} \txt{ such that }  t_\txt{node}(E_{F,2}^{-1}(v))=\txt{`model'} \txt{ and } \txt{deg}^+(v) > 0 \},
	\end{equation}
	where the degree is calculated with respect to $F_2$. This set excludes both the analysis blocks that were unused in the original MCG and those that became unused in step 1 or step 2.

	The set of used global inputs is constructed as
	\begin{equation}
	V_{\txt{in},F} = \Big\{v \in V_\txt{in} \ \Big| \ |\{(x,y) \in E_{F,2}(v) \st y \in V_{A_i} \txt{ for } i \in I_F  \}| > 1 \Big\},
	\end{equation}
	which requires that at least two edges directed out of the nodes in $V_{\txt{in},F}$ must be to the analysis blocks indexed by $I_F$.

	Finally, the set of nodes and edges describing a fundamental problem formulation is given by
	\begin{equation}
	V_F = V_{\txt{in},F} \cup V_\txt{out} \cup \left( \bigcup_{i \in I_F} V_{A_i} \right),
	\end{equation}
	\begin{equation}
	C_{F} = \{ (x,y) \in C_{F,2} \st x,y \in V_F\},
	\end{equation}
	\begin{equation}
	E_F = C_F \cup \left( \bigcup_{i \in I_F} E_{A_i} \right),
	\end{equation}
	\begin{equation}
	F = (V_F,E_F).
	\end{equation}

	\end{enumerate}

>>>>

These operations, though very simple, are effectively the same kind of task 
performed by an analysis block. A constraint or objective function can therfore 
be represented as another analysis block on the graph, with it's own input and 
output variable nodes. Although fundamentally no different than an analysis block, 
for clarity and convience it is useful to distinguish between regular analysis 
blocks and those that arise from the addition of objectives or constraints to 
the graph. Therefore, we define an \emph{expression block}, as the collection of variable and model 
nodes related to a given objective or constraint function. Fig. \ref{f:obj-cons}
highlights the objective and constraint blocks from the Sellar problem.

@@
